# Correlation {#sec-correlation}


```{r}
#| include: false

library(tidyverse)
library(here)

library(stevemisc)
library(patchwork)
```


**Correlation** is a statistical method used to assess a possible **association** between two continuous variables. It is measured by a statistic called the **correlation coefficient**, which represents the **direction** and **strength** of the association between the variables in question.

 

**Range of values**

Correlation coefficient is a **dimensionless** quantity that takes a value in the range **-1** to **+1**.

 

**Direction of association**

A **negative** correlation coefficient indicates that one variable decreases in value as the other variable increases (and vice versa),
a **zero** value indicates that no association exists between the two variables, and a **positive** coefficient indicates that both variables increase (or decrease) in value together.


```{r}
#| echo: false
#| warning: false
#| label: fig-correlation1
#| fig-cap: The direction of association can be (a) negative, (b) no association, or (c) positive.
#| fig-width: 11.8
#| fig-height: 4.0


# plot 1
sigma1<-rbind(c(1,-0.8), c(-0.8,1))
mu1<-c(10, 5) 

set.seed(123)
dat1 <- data.frame(MASS::mvrnorm(n=500, mu=mu1, Sigma=sigma1))

cor1 <- cor(dat1$X1, dat1$X2)

pcor1 <- dat1|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "chartreuse3", alpha = 0.8) +
  ggtitle(paste0("Negative Correlation r=", round(cor1, digits=1))) +
  theme_steve()


# plot2
sigma2<-rbind(c(1,0.000), c(0.000,1))
mu2<-c(10, 5) 

set.seed(124)
dat2 <- data.frame(MASS::mvrnorm(n=500, mu=mu2, Sigma=sigma2))

cor2 <- cor(dat2$X1, dat2$X2)

pcor2 <- dat2|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "grey80") +
  ggtitle(paste0("Zero Correlation r=", round(cor2, digits=1))) +
  theme_steve()


# plot3
sigma3<-rbind(c(1,0.8), c(0.8,1))
mu3<-c(10, 5)

set.seed(125)
dat3 <- data.frame(MASS::mvrnorm(n=500, mu=mu3, Sigma=sigma3))

cor3 <- cor(dat3$X1, dat3$X2)

pcor3 <- dat3|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "deeppink", alpha = 0.8) +
  ggtitle(paste0("Positive Correlation r=", round(cor3, digits=1))) +
  theme_steve()


pcor1 + pcor2 + pcor3 + plot_annotation(tag_levels = 'a')
```


 

**Strength of association**

The **strength** of association can be anywhere between -1 and +1. The **stronger** the correlation, the **closer** the correlation coefficient comes to ±1 (@fig-correlation2). A correlation coefficient of **-1** or **+1** indicates a **perfect** negative or positive association, respectively (@fig-correlation2 c and f). 



```{r}
#| echo: false
#| warning: false
#| label: fig-correlation2
#| fig-cap: The **stronger** the correlation, the **closer** the correlation coefficient comes to ±1.
#| fig-width: 11.8
#| fig-height: 8.0


# plot 4
sigma4<-rbind(c(1,1), c(1,1))
mu4<-c(10, 5) 

set.seed(136)
dat4 <- data.frame(MASS::mvrnorm(n=500, mu=mu4, Sigma=sigma4))

cor4 <- cor(dat4$X1, dat4$X2)

pcor4 <- dat4|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "deeppink") +
  ggtitle(paste0("r=", round(cor4, digits=1))) +
  theme_steve()


# plot5
sigma5<-rbind(c(1,0.8), c(0.8,1))
mu5<-c(10, 5) 
set.seed(139)
dat5 <- data.frame(MASS::mvrnorm(n=500, mu=mu5, Sigma=sigma5))

cor5 <- cor(dat5$X1, dat5$X2)

pcor5 <- dat5|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "deeppink", alpha = 0.8) +
  ggtitle(paste0("r=", round(cor5, digits=1))) +
  theme_steve()




# plot6
sigma6<-rbind(c(1,0.6), c(0.6,1))
mu6<-c(10, 5) 
set.seed(138)
dat6 <- data.frame(MASS::mvrnorm(n=500, mu=mu6, Sigma=sigma6))

cor6 <- cor(dat6$X1, dat6$X2)

pcor6 <- dat6|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "deeppink", alpha = 0.6) +
  ggtitle(paste0("r=", round(cor6, digits=1))) +
  theme_steve()



# plot 7
sigma7<-rbind(c(1,-1), c(-1,1))
mu7<-c(10, 5) 

set.seed(146)
dat7 <- data.frame(MASS::mvrnorm(n=500, mu=mu7, Sigma=sigma7))

cor7 <- cor(dat7$X1, dat7$X2)

pcor7 <- dat7|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "chartreuse3") +
  ggtitle(paste0("r=", round(cor7, digits=1))) +
  theme_steve()


# plot8
sigma8<-rbind(c(1,-0.8), c(-0.8,1))
mu8<-c(10, 5) 
set.seed(149)
dat8 <- data.frame(MASS::mvrnorm(n=500, mu=mu8, Sigma=sigma8))

cor8 <- cor(dat8$X1, dat8$X2)

pcor8 <- dat8|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "chartreuse3", alpha = 0.8) +
  ggtitle(paste0("r=", round(cor8, digits=1))) +
  theme_steve()


# plot9
sigma9<-rbind(c(1,-0.6), c(-0.6,1))
mu9<-c(10, 5) 
set.seed(148)
dat9 <- data.frame(MASS::mvrnorm(n=500, mu=mu9, Sigma=sigma9))

cor9 <- cor(dat9$X1, dat9$X2)

pcor9 <- dat9|>
  ggplot(aes(X1, X2)) +
  geom_point(color = "chartreuse3", alpha = 0.6) +
  ggtitle(paste0("r=", round(cor9, digits=1))) +
  theme_steve()


(pcor9 + pcor8 + pcor7) / (pcor6 + pcor5 + pcor4)   + plot_annotation(tag_levels = 'a')
```


 






